Informing policy via dynamic models: Cholera in Haiti

### 2024-02-26 r1/response.tex

  The authors state they ``develop data analysis strategies leading to improved statistical fit." It's not clear to me exactly what novel strategies have been developed, it seems the authors exclusively use pre-existing techniques from the literature. This is fine, if the focus of the paper is either a) pedagogical or b) on understanding a specific disease system (cholera in Haiti). If any methods are novel, this should be more clearly stated.
}

Fitting models 1 and 2 does not require any methodological innovation. Model 3 has a stochastic spatiotemporal partially-observed structure that leads to methodological challenges. \cite{lee20} made some approximations to avoid carrying out inference for the full coupled system. We show some consequences of those approximations, taking advantage of a newly developed method which has not yet been applied in the context of scientific investigations. So, although the method is not entirely new, its applicability to this task is novel.
Additional commentary to this point has been added in the abstract, introduction, and the subsection on fitting Model 3.

We have further modified the statement in the abstract that led to this confusion. Rather than stating that we ``develop data analysis strategies leading to improved statistical fit", we now say:

\article{We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti. We consider three dynamic models developed by expert teams to advise on vaccination policies. We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit. Speficially, we present appraoches for diagnosing model misspecification and the consequent development of improved models. We futher demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}

@@

1.  To make clear that the issues being addressed by this paper go beyond one modelling study of Haiti, the authors should provide examples and references to support the (accurate in my opinion) statement about, ``common modeling decisions that may not provide an adequate statistical explanation of the data".
}

That statement arose in the author summary, where references are not expected.
However, we appreciate that the point should be expanded on, and we have done so in the revision. Specifically, we cite papers from the literature review we performed on the use of dynamic models for analyzing the Haiti cholera oubreak. The literature review was previously mentioned, but citations from the review were not made. Furthermore, we modified the quoted statement in the author summary to be the following:

\article{we use an in-depth case study to motivate general procedures for fitting mathematical models to time series data, while explaining how these procedures help to address limitations of previous approaches.}

Among the most common mistakes in mechanistic modeling, in our opinion, is not critically evaluating the statistical fit of the model.
We originally tried to make this point by having an entire section dedicated to model diagnostics, though we admit that the connection between the statement in the author summary and a section later in the article was not clear.
We have addressed this by adding sign-posts to references and suggestions back to this statement, and other similar statements in the abtract / introduction.
Our modified discussion also includes detailed examples of principles that were very useful in our case study but are not yet universal practice.

@@

\report{2.
  Given the focus of the paper on how to better perform inference, explicitly commenting on what are the ``existing guidelines for creating models to inform policy [4, 5]". It would then also be helpful in the discussion to comment on how the recommendations of this study extend these pre-existing guidelines.
}

We have added two paragraphs in the introduction discussing how recommendations we make in this article fit in the broader category of papers that make recommendations on fitting mechanistic models to dynamic systems. This includes discussing these two papers that we originally cited in more detail, as well as other articles on similar topics:

\jwc{I combined the two paragraphs into a single paragraph that I think addresses Ed's concern of "brevity and clarity". This paragraph is still quite long and could use some editing, so revisit this later.}

\article{\editModelingGuidance}

% \article{\indent\editModelingGuidanceSecond}

@@

\report{7.
  Usually, vaccine efficacy = 1 implies the vaccine completely blocks transmission and 0 implies no effect. In model 2 the definition of vaccine efficacy (lines 200-205 and Eq 14) implies the opposite. I suggest re-parameterising the model, or renaming the variable. For what its worth, models 1 and 3 seem to follow the usual convention.
}

This change has been made.

\report{8.
  I found the commentary on comparing fitted mechanistic models to a (statistical) benchmark model to be informative and helpful. I was wondering if the authors had any comments on whether (or not) to directly compare the AIC values of the different fitted models directly. Am I right in thinking this should be possible for models 2 and 3 given (I think) they are fitted to the same data? Model 1 is fitted to a different data set, which I understand makes such a comparison impossible.
}

It is correct that models fit to the same datasets can be directly compared using the AIC values. Models~2 and 3 are not fit to the same datasets, however, due to the way Model~3 is initialized. 
Details on how they are initialized are provided in the section with the model descriptions, as well as supplement~S5. 
% To initialize the latent states of Model~3, Lee et al~\cite{lee20} used approximately $60\%$ of the available data as ``burn-in" observations to initialize latent states. According to their supplement, this was done primarily because they believed the dynamics from the start of the outbreak were significantly different enough from the dynamics at the end of the outbreak to suggest that two seperate models would be needed to appropriately describe the data.
% We feel that a properly designed and calibrated model should be able to explain the entire dynamics of the outbreak, including the change from what Lee et al~\cite{lee20} referred to as the ``epidemic" and ``endemic" periods of the outbreak.

% One possible approach to re-analyzing their model would then be to fit all available observations; this would make Models~2 and 3 directly comparable, but would require a significant change to Model~3 itself, as the initialization model used by Lee et al~\cite{lee20} requires a ``burn-in" period. One of our goals, however, was to demonstrate that the models provided by Lee et al.~\cite{lee20} were not the issue, and a complete refactorization of the initialization could potentially contradict this objective. Therefore instead we decided to substantially shrink the ``burn-in" period to only the first three available observations, which corresponds to a natural change in the observations themselves \cite{barzilay13}. These details are available in the Initial Values section in the supplement material.

% In short, though models that are fit to the same dataset are directly comparable using AIC, this is not the case for Models~2 and 3, as the first few observations in the dataset were not considered likelihood evalution of Model~3. This also explains the slight discrepency in AIC for the benchmark models of these two models (Table 2). 
We added the following sentence to clarify this point:

\article{Models that are fit to the same datasets can be directly compared using AIC values, making it a useful tool to compare to benchmark models. Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states, which is described in detail in Sec.~S5 of the supplement.}

@@

\report{9.
  Lines 308-309: Isn't this a log-linear trend in the transmission rate?
}

Yes, that is correct. This has been fixed in the revision. In other instances where the linear trend was previously mentioned, we simply refer to it now as a trend in transmision rate, except where the additional detail of the trend being log-linear is needed for clarity.

@@

Thank you for this suggestion, a sentence on the overall effect of the parameter estimate has been included.
For reference, the maxmum likelihood estimate of the parameter corresponds with a 7.3\% reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval $(1.8\%, 17.9\%)$ for the overall reduction in transmission.

\report{11.
  I might have misunderstood it slightly, but I don't entirely follow the reasoning of lines 340-346. It seems that two different claims are being made: A) the model can’t identify which mechanism underlies a trend in $\beta(t)$ and B) the model can't definitively state there is a trend in $\beta(t)$ compared with some other time-varying parameter (e.g. reporting probability). For model 1, I agree with the authors that claims A and B are valid. What I take issue with is the final statement, which seems to be stronger and apply to both claims and mechanistic models more generally: ``we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism." I agree that the ``incidence data themselves" (which I take to imply ``without additional covariates data") make claim A valid regardless of the model. However, I don’t think claim B holds regardless of the model. By changing model 1 to include time variation in the reporting rate I don't see \emph{a priori} why the ``the incidence data themselves" might not provide enough information to distinguish between time variation in transmission compared with time variation in reporting without the need for additional data. The deficiency is in model 1 rather than the information content of the data. I suggest the authors rework the paragraph to make the reasoning clearer.
}

Our views align with the referee's comments, and we have rewritten the article to make this clearer. The revised paragraphs with this statement are now:

\article{If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
As for any analysis of observational data, we must be alert to the possibility of confounding.
For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled \new{phenomenon} or unmeasured covariate.}

@@

\report{12.
  Lines 553-556: I push back against this point. The models assume an exponential distribution for vaccine-derived immunity. Assuming a mean duration of 10 years, the proportion of individuals who remain immune 9 years after vaccination is . Even for much larger durations of immunity a non-negligible fraction of the population will lose immunity after 9 years, e.g. $e^{-9/10}\approx 0.41$. I therefore don't expect values for the duration of immunity around 10 years to ``effectively result in the same model dynamics".
}

The qualitative comparison we had in mind was the regime with short immunity where many people become infected multiple times during the ten year epidemic, allowing multiple epidemic peaks due to susceptible replenishment.
We have reworded the sentence to avoid making too strong a claim about 10yr immunity being a precise point where the dynamics transition to a regime where reinfection events become negligible.
The revised paragraph is provided below.

\article{Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
Under the framework of Model~2, the best estimate for this parameter is $1.4 \times 10^{11} \, \mathrm{yr}$, suggesting that individuals have permanent immunity to cholera once infected.
Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligble part of the dynamics.
The depletion of susceptible individuals may be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
Perhaps the best interpretation of the estimated parameter, then, is that under the modeling framework that was used, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~2) cautions us against drawing quantitative conclusions from this model.
A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.}

@@

\report{
\textbf{\underline{Major Concerns}}
}

\report{
  \textbf{\underline{1. Literature}}

While the arguments provided throughout the article are well-articulated, there needs to be more supporting literature at the beginning of major sections. For instance, I don’t need to be convinced that the structure of a model should be based on a realistic theory about the observed phenomenon; namely, models should be a white box. However, not everyone is on board with this premise, and supporting literature that argues in favour of this approach should be mentioned. In short, more citations should be added at the beginning of each major section.
}

The referee's suggestions have led to the addition of various references in the revision.
For the particular issue of black box versus white box models, we have included the following paragraph:

@@

\report{
\textbf{\underline{2. Limitations}}

In various passages of this paper, it is hinted that modellers should refrain from deterministic models and instead opt for more realistic stochastic representations. While the critique of ODE models is valid, the shift to stochastic structures is not a free choice. For example, introducing extra-demographic variability adds one additional parameter (infinitesimal variance). In ODE models, one extra parameter can lead to unidentifiability, and there's no apparent reason why this would differ in a stochastic version. Moreover, transitioning to stochastic models involves abandoning well-established MCMC algorithms in favour of methods still in development. Therefore, modellers should not assume that more realistic models with additional parameters are necessarily better without proper caution. In my experience, diagnosing unidentifiability is easier in ODE models than in POMP structures. Hence, there is a trade-off between benefits and costs.
}

The referee raises several points in this paragraph, and we consider them one at a time.

Moving from an ODE model to a stochastic model may not involve any additional parameters, if only demographic stochasticity is included.
We agree, however, that adding to model complexity has costs, including loss of identifiability.
In our revised manuscript, we address identifiability via profile likelihood methods.
% AIC and other formal statistical methods can assess whether the additional complexity is justified by a sufficiently large improvement of statistical fit.
% Identifiability and model simplicity have practical scientific value which goes beyond their statistical role in reducing generalization error and hence improving out-of-fit predictions.
% This is a variation on the issue of ``statistical significance versus practical significance"  often covered in introductory statistics textbooks.

In our study, we find that the deterministic model (model 2) has poor statistical fit.
A poor fit can lead to a risk of identifying spurious scientific phenomena: additional deterministic complexity added to the model may help to fit the model to unmodeled stochastic phenomena present in the data, and may therefore spuriously be assessed as a highly statistically significant model improvement.
We see some evidence of this in our analysis.
Aside from not being stochastic, the mechanistic structure of Model~2 is far more complex than the other two models considered in our study.
Model~2 also demonstrates the most clear cases of spurious scientific findings, as some estimated parameters favor regions of the parameter space that we would not expect scientifically.
This may partially explain the choice of Lee et al~\cite{lee20} to consider a scientifically unjustifiable jump in latent state values from their epidemic to endemic periods: without this additional flexibility, Model~2 cannot adequately describe cholera dynamics in Haiti while keeping parameter values in scientifically plausible ranges.
From our experience, this type of a result is quite common: for a deterministic model to adequately account for unmodeled features of a dynamic system, parameters that are estimated are often forced to unexpected regions of the parameter space.

Adding stochasticity to the model helps address this issue by capturing unmodeled features in the inherent stochasticity of the model rather than in estimated parameters.
Stochasticity in the model alone may not fully solve this problem as evidence suggests that the failure to account for extra-demographic variablility may lead to biases in parameter estimates \cite{he10}.
Our revision includes the addition of profile likelihood for model parameters, which also provide strong evidence for the inclusion of extra-demographic stochasticity in Models~1 and 3 (see supplement material section S8).
Generally, it is advisable to consider the possibility of over-dispersed variation on the transmission rate \cite{breto09,he10} leading to one extra parameter compared to a deterministic model.
In the context of cholera dynamics, \cite{lemaitre19} found a substantial improvement in model fit by including stochasticity, involving one over-dispersion parameter.
\cite{stocks20} found similar results, favoring a stochastic dynamic model with over-dispersion, when investigating transmission dynamics of a different enteric pathogen, rotavirus.
In other contexts, \cite{king15} noted various scientific advantages for including stochasticity in the model; it is undoubtedly part of the biological dynamics, and its omission can lead to over-confident predictions.
Our revised manuscript includes several references demonstrating the importance of modeling extra-demonstraphic stochasticity, including our revised discussion:

We acknowledge the existence of useful mathematical tools for studying ODE models, their identifiability and stability and limiting properties.
Perhaps one example demonstrating the usefulness of ODE models is the Lee et al.~\cite{lee20} study: of the four models considered, the forecasts from Model~2 were most consistent with the actual cholera situation in Haiti.
Despite their inability to describe incidence data, deterministic models can do a good job at detecting overal trends, like the gradual decline of cholera cases in Haiti.
This, coupled with the ease in fitting ODEs to data, makes ODEs a useful tool to gain preliminary insight on a dynamic system.
Furthermore, stochastic dynamic models can be reduced to a deterministic model (a so-called determinsitic skeleton) which can usefully be analyzed to help understand the stochastic dynamics \cite{coulson04}.
We therefore encourage the use of ODE analysis tools whether or not a stochastic model is adopted.
Because this approach is tangential to the points we make in our analysis, and it was not done by Lee et al.~\cite{lee20}, we have not followed that approach in the current manuscript.
A full discussion on deterministic vs simple stochastic models with and without over-dispersion is outside the scope of this article. 
However, we do make a few comments on the benefits of including overdispersion in a model in several locations, including the description of Model~1 and the conclusion. 
Because of their usefulness, we do not go as far as to explicitly forbid the use of deterministic models, but instead point out some of their limitations that should be considered. 
Our revised discussion includes a comment that demonstrates this belief:

\article{Despite their limitations, deterministic models can offer valuable insights into dynamic systems; researchers should carefully weigh these limitations when using these models and presenting their results.}

The decision on whether to choose Markov chain Monte Carlo (MCMC) methods is not fully aligned with the decision on whether to include stochasticity in the model.
In our literature review, of the 12 previous Haiti cholera investigations including stochastic dynamic models \cite{kirpich17,lee20,pasetto18,mukandavire13,kirpich16,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}, four of them carried out inference via MCMC \cite{pasetto18,lewnard16,sallah17,azman12}.
We agree with the referee that MCMC is more routine for determinstic models than for stochastic models, and advanced MCMC methods such as Hamiltonian Monte Carlo become more readily applicable.
Sequential Monte Carlo (SMC) was used for only two of the investigations in our literature review \cite{lee20,azman15}.
In principle, MCMC methods can be applied for inference in over-dispersed stochastic dynamic models via so-called particle MCMC (PMCMC) \cite{andrieu10}.
In practice, these models have been fitted by maximum likelihood, some examples including \cite{lemaitre19,fox22,molodecky23,pons-salort18,subramanian20,stocks20}.

@@

The particle filter provides access to the likelihood of the data, and conditional likelihood of each individual data point.
By contrast, obtaining reliable likelihood values from MCMC is not straightforward.
Thus, particle filter methods facilitate the use of various likelihood-based methods for diagnostics and model selection.
We demonstrate this in Figs S4, S5, S6, S7. \TODO{Check all figure numbers, especially to the supplement.}

As a proper scoring rule \cite{gneiting07}, log-likelihood can be used as an effective way to evaluate filters as well as models. 
The particle filter is unbiased for the likelihood, and therefore negatively biased for the log-likelihood with bias approaching zero as the number of particles increases and the Monte Carlo variance decreases.
%%Log-likelihood is a proper scoring rule for probabilistic forecasts \cite{gneiting07}, so any error in filtering is expected to lead to a loss in log-likelihood. Thus, log-likelihood is an appropriate way to evaluate filters as well as models.
If optimization searches from diverse starting points provide a consensus on the maximum log-likelihood, and the Monte Carlo variance is small, we can have reasonable confidence in our maximization.
Monte Carlo adjusted profile (MCAP) methods \cite{ionides17,ning21} provide methodology which reduces and evaluates the consequences of this Monte Carlo error.
MCAP methods have been used in various recent investigations using particle filters \cite{subramanian21,ali22,pons-salort18}.

### 2024-02-26 edits.tex
First round of edits of revision based on Ed's comments

\usepackage[table]{xcolor}
\definecolor{mygreen}{RGB}{0, 90, 0}
\newcommand\original[1]{\textcolor{red}{#1}}
\newcommand\TODO[1]{\textcolor{red}{[TODO: #1]}}

@@

%They recommend five guiding principles: stakeholder engagement, complete model documentation, complete description of data used, communicating uncerainty, and testable model outcomes.
\newcommand\editModelingGuidanceFirst{There are many guidelines available regarding the use of mechanistic models for policy influence. For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards. They used this review to develop general recommendations for policy-guiding modeling. This review is particularly relevant given that their recommendations are specifically intended for the modeling of neglected tropical diseases, which includes cholera. Their resulting recommendations for communicating model results align with other manifestos on the proper use of models intended to inform policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communicating uncertainty. Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on how to effectively model dynamic systems.}

\newcommand\editModelingGuidanceSecond{Based on our assessment and literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling. Concretly, \cite{lee20} follows at least four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}. Determining the level of adherence to the first principle, \emph{stakeholder engagement}, is difficult based solely on the article. Despite this, the inconsistency between their forecasts and the cholera incidence from 2019 to 2022 suggests that existing recommendations and standards related to policy-driven modeling may be insufficient. \cite{saltelli19} suggests that improvements in model model-based outputs may be obtained by developing structures and standards based on statistical principles. Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework. We specifically emphasize principles that proved essential in our case study; complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are given in \cite{li23}.}

@@

\newcommand\editDiscussionTwo{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric like AIC. Nested model variations are particularly useful as they enable formal statistical testing of the nested features. The cholera case study demonstrated the advantage of such a model variation in all three models. For instance, we tested the inclusion of a trend in transmission rate over time in Model~1; in Model~2, we considered adding an additional phase parameter, which enables a shift in the seasonal peaks of cholera infections; in Model~3, we included additional parameters such as those associated with Hurricane Matthew. In all these cases, we found strong statistical evidence in favor of these model variations over their simpler counterparts.}

\newcommand\editDiscussionThree{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model. Mathematical models necessarily simplify complex dynamical systems in order to make inference possible. Decision-making about which features of a system need to be incorporated in a model and which should be excluded can significantly influence the model-based conclusions. However, these decisions can be tested using a likelihood-based framework. Standard model decisions cover deterministic versus stochastic modeling and the decision to include or exclude a meta-population structure. Model benchmarks can assist in determining the level of complexity to be used in the model, enabling researchers to gauge a model's relative performance against a better-understood statistical model.}

\newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the feature greatly impacts observed data. This is because the model must compensate for these unaccounted-for features using its inherent flexibility, particularly in the calibration of parameters. In deterministic models, parameter estimation is the sole source of such flexibility. When the influence of unmodeled dynamics is significant, it can skew parameter estimates toward scientifically implausible values, a phenomenon observed with Model~2 in our case study. Incorporating demographic and environmental stochasticity into models can mitigate the impact of these unmodeled features. These stochastic processes are not only arguably present in most dynamic systems, but their inclusion in a model also allows observed data variations to be attributed to inherent randomness rather than to distorted parameter values. Furthermore, incorporating extra-demographic stochasticity in a model has also been shown to be beneficial \cite{he10,stocks20,li23}, which aligns with our findings: Models~1 and 3 suggest the presence of extra-demographic stochsticity, as evidenced by the confidence intervals provided in Suplement S8.}

\newcommand\editDiscussionFive{The three models in this case study offer a unique combination of decisions of model complexity. Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences. In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm. One of only a few examples of fitting a model with this level of complexity, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}

\newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance. This is particularly true for long time series, as the latent states from simulations from initial conditions may be more likely to diverge from the truth as time increases. One way to address this issue is by simulating forward from the filtering distribution at the last available time point, as this proceedure considers available evidence about the latent states of a system under the framework of the dynamic model. This type of forecasting, however, is not possible using a deterministic model. Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}. Despite these limitations, forecasts from deterministic models can provide valuable insights on potential future trends, though researchers should be transparent about these limitations when presenting forecasts from deterministic models.}

@@

\newcommand\editForecastOne{Forecasts of a dynamic system should ideally be based on the most recent information available as it is likely to be more relevant than older data. While this assertion may seem self-evident, it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance. Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}. Lee et al.~\cite{lee20} chose to obtain forecasts from all of their models by simulating forward from initial conditions.
This decision is possibly as a result of using a deterministic model: forecasts from different models may only be considered comparable only if they are obtained in the same way, which must be by simulating from initial conditions because Model~2 is deterministic.}

\newcommand\editForecastTwo{In contrast, for non-deterministic Models~1 and 3, we obtain forecasts by simulating future values using latent states that are harmonious with the most recent data. This is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$. The decision to obtain model forecasts from initial conditions partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}. Table~S7 in their supplement material, which contains results that were not discussed in their main article, shows that the subset of their simulations which were consistent with observing zero cases from 2019-2020 were also more consistent with the disappearance of cholera from Haiti from 2019-2022. These results support our argument: forecasts of a dynamic system should start with latent states that are consistent with available data.}

@@

>>>>

\usepackage[table]{xcolor}
\definecolor{mygreen}{RGB}{0, 90, 0}
% \definecolor{purple}{RGB}{128,0,128}
\newcommand\original[1]{\textcolor{red}{#1}}
\newcommand\TODO[1]{\textcolor{red}{[TODO: #1]}}

@@

%They recommend five guiding principles: stakeholder engagement, complete model documentation, complete description of data used, communicating uncerainty, and testable model outcomes.
% \newcommand\editModelingGuidanceFirst{There are many guidelines available regarding the use of mechanistic models for policy influence. For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards. They used this review to develop general recommendations for policy-guiding modeling. This review is particularly relevant given that their recommendations are specifically intended for the modeling of neglected tropical diseases, which includes cholera. Their resulting recommendations for communicating model results align with other manifestos on the proper use of models intended to inform policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communicating uncertainty. Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on how to effectively model dynamic systems.}

\newcommand\editModelingGuidance{There are many guidelines available regarding the use of mechanistic models for policy influence. For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards. Their recommendations generally align with other manifestos on the proper use of models intended to inform policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communicating uncertainty. Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on how to effectively model dynamic systems.  Based on our assessment and literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling. Concretly, \cite{lee20} follows at least four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}. Determining the level of adherence to the first principle, \emph{stakeholder engagement}, is difficult based solely on the article. Despite this, the inconsistency between their forecasts and the cholera incidence from 2019 to 2022 suggests that existing recommendations and standards related to policy-driven modeling may be insufficient. \cite{saltelli19} suggests that improvements in model model-based outputs may be obtained by developing structures and standards based on statistical principles. Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework. We specifically emphasize principles that proved essential in our case study; complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are given in \cite{li23}.}

@@

\newcommand\editDiscussionTwo{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric such as AIC. Nested model variations are particularly useful as they enable formal statistical testing of the nested features. The cholera case study demonstrated the advantage of such a model variation in all three models. For instance, we tested the inclusion of a trend in transmission rate over time in Model~1; in Model~2, we considered adding an additional phase parameter, which enables a shift in the seasonal peaks of cholera infections; in Model~3, we included additional parameters such as those associated with Hurricane Matthew. In all these cases, we found strong statistical evidence in favor of these model variations over their simpler counterparts.}

\newcommand\editDiscussionThree{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model. Mathematical models necessarily simplify complex dynamical systems in order to make inference possible. Decisions about which features of a system to incorporate in a model and which should be excluded can significantly influence the model-based conclusions. However, these decisions can be tested using a likelihood-based framework. Standard model decisions cover deterministic versus stochastic modeling and the decision to include or exclude a meta-population structure. Model benchmarks can assist in determining the level of complexity to be used in the model, enabling researchers to gauge a model's relative performance against a better-understood statistical model.}

\newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the feature greatly impacts observed data. This is because the fitted model must compensate for these unaccounted-for features using its inherent flexibility, particularly in the calibration of parameters. In deterministic models, parameter estimation is the sole source of such flexibility. When the influence of unmodeled dynamics is substantial, it can skew parameter estimates toward scientifically implausible values, a phenomenon observed with Model~2 in our case study. Incorporating demographic and environmental stochasticity into models can mitigate the impact of these unmodeled features. These stochastic phenomena are not only arguably present in most biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent randomness rather than to distorted parameter values. Furthermore, incorporating extra-demographic stochasticity in a model has also been shown to be beneficial \cite{he10,stocks20,li23}, which is consistent with our results: Models~1 and 3 suggest the presence of extra-demographic stochsticity, as evidenced by the confidence intervals provided in Suplement S8.}

\newcommand\editDiscussionFive{The three models in this case study consider a range of possible model complexity. Of these models, Model~3 is distinct in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences. In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm. One of only a few examples of fitting a model with this level of complexity via maximum likelihood, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}

\newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance. This is particularly true for long time series, as the latent states from simulations from initial conditions may be more likely to diverge from the truth as time increases. One way to address this issue is by simulating forward from the filtering distribution at the last available time point, as this proceedure considers available evidence about the latent states of a system under the framework of the dynamic model. This type of forecasting, however, is not possible using a deterministic model. Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}. Despite their limitations, deterministic models can offer valuable insights into dynamic systems; researchers should carefully weigh these limitations when using these models and presenting their results.}

@@

\newcommand\editForecastOne{Forecasts of a dynamic system should should be consistent with the available data. It is particularly important that forecasts are consistent with the most recent information available, as recent data is likely to be more relevant than older data. While this assertion may seem self-evident, it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data may not be consistent with model trajectories. Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}. Lee et al.~\cite{lee20} chose to obtain forecasts from all of their models by simulating forward from initial conditions, rather than conditioning forecasts based on the available data.
This decision is possibly as a result of using a deterministic model, as forecasts from different models may only be considered comparable if they are obtained in the same way, which must be by simulating from initial conditions because Model~2 is deterministic.}

\newcommand\editForecastTwo{In contrast, for non-deterministic Models~1 and 3, we obtain forecasts by simulating future values using latent states that are harmonious with the most recent data. This is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$. The decision to obtain model forecasts from initial conditions partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}. Table~S7 in their supplement material, which contains results that were not discussed in their main article, shows that the subset of their simulations which were consistent with observing zero cases from 2019-2020 were also more consistent with the disappearance of cholera from Haiti from 2019-2022. These results support our argument that forecasts should be made by ensuring the starting point for the forecast is consistent with available data.}

@@

\newcommand\editSpatBench{To obtain a benchmark for models with a meta-population structure, we fit independent auto-regressive negative binomial models to each spatial unit. Under the assumption of independence, the log-likelihood of the benchmark on the entire collection of data can be obtained by summing up the log-likelihood for each independent model. In general, a spatially explicit model may not have well-defined individual log-likelihoods, and, in this case, comparisons to benchmarks must be made at the level of the joint model.}
